---
title: "R Markdown"
subtitle: "Maria C. Ramos"
author: ""
institute: "MiR Study Group"
date: "04/04/2021"
output:
  xaringan::moon_reader:
    seal: FALSE
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, center, middle

# R Markdown

<br><br>.large[Maria C. Ramos| MiR Study Group | 20 Feb 2021]

<br><br><br>`r icon::fa("link")`.large[mariacramos.com]
<br><br>`r icon::fa("twitter")`.large[@mariacramosf]
---
class: top, left
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Outline

* reading files 

* parsing files

* parsing vectors


---
class: middle, center, middle

# Why R Markdown

## 1) Code 
## 2) Results
## 3) Comment 
## 4) 

---

# Reading Files: General Strategy 

### Step 1: select the appropriate package and function according to the format of your data file (reading)

--

### Step 2: Take care of general aspects (column names, metadata, and NAs)

--

### Step 3: Ensure R correctly classifies the data types within the file (parsing)

---
# Reading Files 

### Step 1: select the appropriate package and function according to the format of your data file 

#### Package readr

.center[
```{r, echo=FALSE}
dfn <- data.frame(`readr function` = c("read_csv( )", 
                               "read_csv2( )",
                               "read_tsv( )", 
                               "read_delim( )", 
                               "read_fwf( )",
                               "read_table( )", 
                               "read_log( )"),
                  `file format`=c("comma delimited", "semicolon delimited", "tab delimited", "any delimited", "fixed width", "fixed width delimited by white space", "Apache style files"))
dfn %>%
   knitr::kable("html")
```
]

---
# Reading Files 

### Step 1: select the appropriate package and function according to the format of your data file 

#### Other packages  

.center[
```{r, echo=FALSE}
dfn2 <- data.frame(`package` = c("haven", 
                               "readxl",
                               "DBI"),
                  `data format`=c("SPSS, SAS, Stata", "excel files", "SQL"))
dfn2 %>%
   knitr::kable("html")
```
]

---

# Reading Files 

### Step 2: Take care of general aspects (column names, metadata, and NAs)

#### Focus on most commonly used function `read_csv`

---

by default, `read_csv()` reads first line as column names

## Problem focused strategies 

.center[
```{r, echo=FALSE}
dfn3 <- data.frame(problem = c("n first rows have metadata", "observations that start with '#' are actually comments","first row does not contain the column names", "different coding of NA"), strategy =c("use argument skip = n to skip first n rows", "use argument comment = to skip lines that start with a symbol", "use argument col_names = FALSE to avoid treating first row as column names or set col_names = c(set of col names) to manually set column names", "use argument na = to specify which values to treat as NA"))
dfn3 %>%
   knitr::kable("html")
```
]

---

# Reading Files 

### Step 3: Ensure R correctly classifies the data types within the file (parsing) 

#### How parsing works 

* readr reads the first 1000 rows and parses the data type of each column.
* If no other match is found, columns are parsed as character.

---

## Common file parsing problems

* first 1000 rows might not be representative
* there are too many NAs

## General file parsing strategies 

* column by column assess parsing and correct (if needed)
* use `problems()` to see parsing problems
* use argument `col_types = cols()` to specify the correct parsing for each column


```{r, eval=FALSE}
myfile <- read_csv("filepath.csv",
                   col_types = cols(#<<
                     x = col_double(), #<<
                     y = col_date(), #<<
                     z = col_character()#<<
                   ))
```

---

### Other file parsing tips

* can parse every column as character and go from there

```{r, eval=FALSE}
myfile <- read_csv("filepath.csv", 
           col_types = cols(#<<
             .default = col_character()#<<
           ))
```

* If working with very large file, use argument `n_max` to increase the number of rows used to parse. 

---
# Parsing vectors 

### Parsing a file versus parsing vectors 

`parse_*()` takes a character vector and transforms it into a vector of another type. 

.center[
```{r, echo=FALSE}
dfn4 <- data.frame(files = c("get data into R", "uses col_* argument" ), vectors =c("data already in R", "uses parse_* functions"))
dfn4 %>%
   knitr::kable("html")
```
]

---

# Parsing Functions

.center[
```{r, echo=FALSE}
dfn5 <- data.frame(functions = c("parse_double()", "parse_number()", "parse_logical()", "parse_character()", "parse_integer()", "parse_datetime()", "parse_date()", "parse_time()"))
dfn5 %>%
   knitr::kable("html")
```
]

---

# When Parsing Fails...

* You will get a warning with first failures

* Parsing failures will be missing from the output

* Use `problems()` to get complete list of failures

---

# Parsing Numbers

.center[
```{r, echo=FALSE}
dfn6 <- data.frame(issues = c("decimal marks vary", "some text is surrounding the number (e.g., $, USD, %)", "grouping marks vary"), solutions =c("use parse_double(vector, locale = locale(decimal_mark = ))", "use parse_number -> automatically removes surrounding text", " use parse_number with argument locale = locale(grouping_mark = )"))
dfn6 %>%
   knitr::kable("html")
```
]

.pull-left[
**Decimal Mark**
```{r, eval=FALSE}
numeric_vector <- 
  parse_double(vector,
               locale = 
                 locale(
                   decimal_mark = ".")))#<<
```
]

.pull-right[
**Grouping Mark** 
```{r, eval=FALSE}
numeric_vector <- 
  parse_number(vector, 
               locale = 
                 locale(
                   grouping_mark = ",")))#<<
```
]

---
# Parsing Characters 

**Key issue:** encoding

* Look at your data. If you find weird characters, R did not correctly detect the encoding

### Solutions

.center[
```{r, echo=FALSE}
dfn7 <- data.frame(situation = c("if you know then encoding", "if you don't know the encoding"), solutions =c("use argument locale = locale(encoding = )", "use guess_encoding(charToRaw(vector or file name)). Then try the encodings with argument locale in cell above"))
dfn7 %>%
   knitr::kable("html")
```
]

---

# Solutions to Encoding Issues

.pull-left[
**You know the encoding**
```{r, eval=FALSE}
character_vector <- 
  parse_character(vector,
               locale = 
                 locale(
                   encoding = "Latin1")))#<<
```
]

.pull-right[
**You don't know the encoding** 
```{r, eval=FALSE}
guess_encoding(charToRaw(vector_name))
```
]


---

# Parsing Factors 

**Key issue:** inconsistent names for levels

**Detection:** check the levels of your vector after parsing

**Solution:** provide level names with `levels =` argument 

---
 
# Parsing Dates and Times

**Key issue:** Data format doesn't match format parsing function expects

**Solution:** Use argument `format =` inside the parsing function

.center[
```{r, echo=FALSE}
dfn8 <- data.frame(functions = c("parse_datetime()", "parse_date()", "parse_time()"),  expected_format = c("year, month, day in descending order", "1999-02-27 or 1999/02/27", "23:20:20am - min, sec, and am/pm are optional"))
dfn8 %>%
   knitr::kable("html")
```
]


---
class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).
